---
output: md_document
---

```{r echo=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  comment = ">"
)
```

# nicheApport

#### Tokeshi's Niche Apportionment Species Abundance Distributions Models

**Mario J. Marques-Azevedo**

## Description
`nicheApport` package simulate and fit the stochastic niche-oriented species abundance distribution models proposed by Tokeshi (1990, 1996). This package simulate species rank abundance distributions (RADs) and use Monte Carlo approach, proposed by Bersier & Sugihara (1997) and improved by Cassey & King (2001) and Mouillot et al. (2003), to fit models to replicated data.

The nicheApport package provides:

  * Simulate species abundance models: Dominance Decay, Dominance Preemption, MacArthur Fraction, Random Fraction, Random Assortment and Power Fraction;
  
  * Fit simulated rank abundance distribution to observed rank abundance data.

### References
Bersier, L.-F. & Sugihara, G. 1997. Species abundance patterns: the problem of testing stochastic models. *J. Anim. Ecol.* 66: 769-774.

Cassey, P. & King, R. A. R. 2001. The problem of testing the goodness-of-fit of stochastic resource apportionment models. *Environmetrics* 12: 691-698.

Mouillot, D. et al. 2003. How parasites divide resources: a test of the niche apportionment hypothesis. *J. Anim. Ecol.* 72: 757-764.

Tokeshi, M. 1990. Niche apportionment or random assortment: species abundance patterns revisited. *J. Anim. Ecol.* 59: 1129-1146.

Tokeshi, M. 1996. Power fraction: a new explanation of relative abundance patterns in species-rich assemblages. *Oikos* 75: 543-550.

## Installation

### Stable version

```{r, eval = FALSE}
library(devtools)
install_github(repo = "mariojose/nicheApport", ref = "master", build_vignettes = TRUE)
```

### Development version
This version have the last updates waiting to release.

```{r, eval = FALSE}
library(devtools)
install_github(repo = "mariojose/nicheApport", ref = "dev", build_vignettes = TRUE)
```

Note that if you are running Microsoft Windows, you need to install the [`Rtools`](https://cran.r-project.org/bin/windows/Rtools/index.html) package to build vignettes locally. If you do not want build vignettes, remove 'build_vignettes = TRUE' argument.

## Example of usage

We can simulate one of the six models informing number of species, total abundance and type of data:

```{r models_examples}
# Load library
library(nicheApport)

# Dominance Decay model with 10 species and 100 individuals
# ('count = TRUE' define that data is discrete)
dominanceDecay(N = 100, S = 10, count = TRUE)

# Dominance Preemption model with 10 species and 100 kg,
# for instance, as total abundance.
# ('count = FALSE' define that data is continuous - default value)
dominancePreemp(N = 100, S = 10, count = FALSE)

# Power Fraction model with 10 species, 100 individuals and
# 'w' parameter iqual 0.2
powerFraction(N = 100, S = 10, w = 0.2, TRUE)
```

For fitting model to data we can use _fitmodel_ function or _fitmodelCl_ function to run simulation in cluster mode.

```{r fit_example}
# Creating an abstract data with 10 replicates and 40 ranks (species)
dt <- matrix(nrow = 10, ncol = 40)
for(i in 1:length(dt[ ,1])){
  dt[i, ] <- randFraction(N = 100, S = 40, count = FALSE)
}

# Fitting model to data
m_rf<- fitmodel(x = dt, model = "randFraction", count = FALSE, nRand = 99)

# Printing statistics for fitting
m_rf

# Get simulated data statistics
m_rf@sim.stats
```

## Contribution and citation

The _nicheApport_ project are hosted in [GitHub](https://github.com/mariojose/nicheApport/) and [issues or bugs report](https://github.com/mariojose/nicheApport/issues) are welcome.

For citing _nicheApport_ use _citation(package = 'nicheApport')_ function in R.

